---
title: "W271 Section 3 Lab 2"
author: "Kiersten Henderson, Jill Zhang, Hoang Phan, Daghan Altas"
date: "10/22/2017"
output: pdf_document
---

```{r setup, warning=FALSE, message=FALSE, size="tiny"}
library(easypackages)
packages("knitr","Hmisc","ggplot2","tidyverse","forcats", "gridExtra","reshape", "stargazer", "ordinal","car","plyr")
opts_chunk$set(tidy.opts=list(width.cutoff=75),tidy=TRUE)
```

#1 Introduction  

Our team of data scientists is working with a University foundation to identify alumni who are likely to donate in the future. We were able to model not just whether alumni with particular characteristics would donate, but rather what level of contribution particular alumni were most likely to make. In order to evaluate what variables to include when we estimated our model, we considered fundraising domain knowledge and the results of our explanatory data analysis (EDA).

For several reasons supported by our EDA (below), we modeled donation level as a categorical variable and used a proportional odds model to estimate level of donation as the dependent variable. 

1. We elected to use an Ordinal model as the donation brackets are clearly ordered.
2. After completing multiple levels of EDA, we found key trends and nuances in the data that we used for modeling
3. We compared 3 models on their alignment with our EDA, goodness of fit, and logical real-world structure, and ultimately chose a final model that exceled in all three of these aspects.

The final model we estimated is:

$$ logit(\hat{P}(Y\leq j)) = \hat{\beta}_{j0} + 0.07Married + 0.27Attended + 1.21FY15Giving \ ... + 2.10FY12Giving.Grouped $$

where $\hat{beta}_{10} = 2.73, \hat{beta}_{20} = 4.38, \hat{beta}_{30} = 6.82, \hat{beta}_{40} = 8.21$

This model primarily consists of previous years' donations, as we found they were the best predictors of donating in the current year. Fundraising professionals typically consider an individual's wealth (ability to give) and enthusiasm for giving when they consider who might be the most likely, loyal, and generous donors to a particular cause. For this reason, we included Advanced Degree as a potential indicator of wealth (those individuals with graduate degrees frequently have higher incomes). In addition, we grouped majors together according to level of donations typically made by donors from each major. We also considered enthusiasm for and support of the university which was evident if an alumnus had attended a university event, thus we included Event Attendence as well. In addition, we found in our EDA that Gender and Marital Status contributed to identifying the philanthropic behavior towards the University and we thus included Gender and Marital status when we estimated our model.


#2 Exploratory Data Analysis

```{r}
givings = read.csv("./lab2data.csv")
str(givings)
sum(is.na(givings))
```
##2.1 Observations  

When we conducted a cursory examination, we found that there are 1000 observations and twelve variables in the dataset (five variables are donations in years 2012-2016). There were no missing values for any of the variables in the dataset. 

Donations made in the 2016 fiscal year is our dependent variable. We were provided donations for 5 years as continuous variables. The minimum donation is $0 and the maximum donation is $161 500 (made in 2013).
We were provided Gender and Attendence.Event as a binary variables. Furthermore, we were provided Marital.Status as a factor with four levels (D, M, S, W), which we interpret as divorced, married, single, and windowed. Our data set includes graduating class year with five categories each ten years apart (1972, 1982, 1992, 2002, 2012). We were also provided with educational information for alumni; Alumni Major was provides in 45 categories, and information about their next degree was provided with 47 categories.

##2.2 Data Cleaning

To facilitate interpretation of categories during EDA and modeling, we renamed some of them. In addition, we assigned categorical variables to factors where appropriate.


```{r}
levels(givings$Gender)  = c("Female", "Male")
givings$AttendenceEvent = factor(givings$AttendenceEvent, levels = c(0,1), 
                                 labels = c("Didn't Attend", "Attended"))
levels(givings$Marital.Status) = c("Divorced", "Married", "Single", "Widowed")
givings$Class.Year = as.factor(givings$Class.Year)
givings$FY12Giving = as.numeric(givings$FY12Giving)
givings$FY13Giving = as.numeric(givings$FY13Giving)
givings$FY14Giving = as.numeric(givings$FY14Giving)
givings$FY15Giving = as.numeric(givings$FY15Giving)
givings$FY16Giving = as.numeric(givings$FY16Giving)
```  


In addition, we created a factor variable from FY16Givings as suggested. We did so by grouping FY2016  donations into 5 buckets and we extended this logic to all other fiscal year donations.

```{r size = 'tiny'}
givings$FY12Giving.Grouped <- factor(cut(givings$FY12Giving, breaks=c(0,1,100,250,500, 200000), labels=c("[0-1)", "[1-100)", "[100-250)","[250-500)","[500-200000)"), right = FALSE))
givings$FY13Giving.Grouped <- factor(cut(givings$FY13Giving, breaks=c(0,1,100,250,500, 200000), labels=c("[0-1)", "[1-100)", "[100-250)","[250-500)","[500-200000)"), right = FALSE))
givings$FY14Giving.Grouped <- factor(cut(givings$FY14Giving, breaks=c(0,1,100,250,500, 200000), labels=c("[0-1)", "[1-100)", "[100-250)","[250-500)","[500-200000)"), right = FALSE))
givings$FY15Giving.Grouped <- factor(cut(givings$FY15Giving, breaks=c(0,1,100,250,500, 200000), labels=c("[0-1)", "[1-100)", "[100-250)","[250-500)","[500-200000)"), right = FALSE))
givings$FY16Giving.Grouped <- factor(cut(givings$FY16Giving, breaks=c(0,1,100,250,500, 200000), labels=c("[0-1)", "[1-100)", "[100-250)","[250-500)","[500-200000)"), right = FALSE))
```


##2.3 Univariate Data Analysis


We conducted univariate analysis of the following variables:

* FY12 though FY16 Giving (numerical, log transformed and Grouped)
* Gender
* Class.Year
* Marital.Status
* Major
* Next.Degree
* AttendenceEvent

###2.3.1 FY12 though FY16 Giving (Numerical, log transformed, and Grouped)

We note that continuous-scale values of the contributions variable have a very strong positive skew. However, at log scale, we observe a bi-modal distribution, with most of the values centered either around 0 or around the $100 range.Therefore, we deem that a log transformation is not an appropriate way to specify the model.

As noted in the data cleaning section, we transformed the all years' giving variables into a categorical variable. We think this approach is justified because all years exhibit a very strong skew and becasue the log transformation was not a good solution and would hinder model interpretation. Binning this variable addressed our concerns and still allows for ease of model interpretation.

From this univariate analysis alone, we can observe that most years follow a similar pattern - ie. donor behavior seems to be consistent over multiple years. We anticipate that there may be a strong correlation between a donor’s 2016 donation level and their donation in previous years and we will explore this idea during bivariate analysis. This will be important to explore because it would be important to uniquely identify and model donors who consistently make large contributions because the University can use this information to maximize yearly total donations.



```{r}
givings.tidy.donations <- givings[1:12] %>% gather("Giving.Year", "Donations", 8:12)
givings.tidy.donations$Giving.Grouped <- factor(cut(givings.tidy.donations$Donations, breaks=c(0,1,100,250,500, 200000), 
                                                    labels=c("[0-1)", "[1-100)", "[100-250)","[250-500)","[500-200000)"), right = FALSE))
givings.tidy.donations.aggregate <- as.data.frame(xtabs(~ Giving.Grouped + Giving.Year, data = givings.tidy.donations))
p1 <- ggplot(givings.tidy.donations, aes(x=Donations, colour=Giving.Year)) +
  geom_density(alpha=0.3) +
  labs(title = "Density plot for contributions") +
  theme_bw()

p2 <- ggplot(givings.tidy.donations, aes(x=Donations, colour=Giving.Year)) +
  geom_density(alpha=0.3) +
  scale_x_continuous(breaks=c(0,1,100,250,500,200000),trans="log1p", expand=c(0,0)) +
  labs(title = "Log scale density plot for contributions") +
  scale_y_continuous() +
  theme_bw()

p3 <- ggplot(givings.tidy.donations.aggregate, aes(x = Giving.Grouped, y=Freq)) + 
  geom_bar(aes(fill = Giving.Year ), stat = 'identity', position = 'dodge') + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y = "Donations", x= "Brackets", title = "Contributions over years") +
  theme_bw()
grid.arrange(p1, p2,p3, ncol=1, nrow=3)
```

###2.3.2 Gender

The dataset contains nearly identical numbers of female and male donors. This result is surprising 
because recently, male and female enrollment is quite different. According to the National Center for Education Statistics, the national average in 2015 was 56% female and 44% for male enrollment in college (https://nces.ed.gov/programs/coe/indicator_cha.asp) and graduation rates are also skewed towards women. However, because college graduation rates were likely skewed toward men in earlier years, they may average out to give an equal distribution of men and women during the period between 1972 and 2012. Furthermore, it is possible the particular University were are studying is unusual in that it has always had equal proportions of male and female alumni.

```{r}
row <- xtabs( ~ Gender, data = givings)
data.frame(rbind(row,row/dim(givings)[1]), row.names = c("Donor Count", "Ratio"))
```


###2.3.3 Class.Year

We were surprised that only 5 graduation years are provided. Because of this, the individuals sampled are not a random subsample from the entire population of university graduates, but rather a subsample of students with each set spaced apart by 10 years. This is a major caveat and we will be cautious when applying our model to all alumni of the University. 

```{r}
row <- xtabs( ~ Class.Year, data = givings)
data.frame(rbind(row,row/dim(givings)[1]), row.names = c("Class.Year Count", "Ratio"))
```

###2.3.4 Marital.Status

We were surprised to find that the Divorce to Marriage ratio is very low compared to the the expected ratio of 44% (Wikipedia https://en.wikipedia.org/wiki/Divorce_demography). That said, the measurement methodology is slightly different and we expect rates to change with graduation years (divorce rates are more likely to increase with age). So we are going to assume that Marital.Status data is valid sample for the population.

```{r}
row <- xtabs( ~ Marital.Status, data = givings)
data.frame(rbind(row,row/dim(givings)[1]), row.names = c("Marital.Status Count", "Ratio"))
```


###2.3.5 Major

Many of the categories of major have very little representation (for example Zoology, Political studies in regional studies) so we don’t expect any one of them to make a significant contribution to our model. When we grouped major into larger categories (Liberal Arts, Sciences, etc - 11 categories in total), we also did not gain any insight into donor contributions. We will, however, examine Majors to identify any outlier majors that may have donated disproportionately and use that information to improve our modeling.

```{r fig.height=1.5, warning = FALSE}
ggplot(givings, aes(x = Major)) + geom_histogram(stat ="count") + labs(title = "Count for each Major", x = "Various Majors") +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())
head(sort(xtabs( ~ Major, data = givings)),3)
```


###2.3.6 Next.Degree

The Next.Degree variable has too many categories to be useful either as-is or grouped into subcategories.Many levels only have a single count (ex: MA2, MALS,MSM, BD, etc). We therefore grouped Next.Degree into 3 categories; alumni without a next degree, and those with next degree being a bachelors or above a bachelors (graduate level or professional degree). 

```{r fig.height= 1.5, warning=FALSE, size="tiny"}
ggplot(givings, aes(x = Next.Degree)) + geom_histogram(stat ="count") + labs(title = "Count for each Next.Degree", x = "Various Degrees") +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())

givings$Adv.Deg <-fct_collapse(givings$Next.Degree,
                               bachelor_equivalent = c("AA","BA","BAE","BD","BFA","BN","BS","BSN","LLB","LLD","NDA","UBDS","UDDS","UMD","UMDS","UNKD","TC"), 
                               above_bachelor= c("DC","DDS","DMD","DO","DO2","DP","JD","PHD","MA","MA2","MAE","MALS","MAT","MBA","MCP","MD","MD2","ME", "MFA","MHA","ML","MLS","MM","MPA","MPH","MS","MSM","MSW","STM"))
```
The Next.Degree as a factor variable is too scathered. Many levels only have a single count (ex: MA2, MALS, MSM, BD, etc). We will group donor into 3 catories;  those without a next degree (None), those with a bachelor equivalent and those with a degree higher than bachelor.  

###2.3.7 AttendenceEvent

Surprisingly, 40% of graduates have attended at least one Alumni event organized between 2012 and 2015. Intuitively, we expect a high correlation between this variable and donations (if alumni support, or are enthusiastic about their, they are more likely to donate) so we included this variable in our analysis and modeling.


```{r}
row <- xtabs( ~ AttendenceEvent, data = givings)
data.frame(rbind(row,row/dim(givings)[1]), row.names = c("AttendenceEvent Count", "Ratio"))
```


##2.4 Bivariate Data Analysis


We are examined the relationship between the following sets of variables:

* FY16Giving.Grouped vs. (Gender, Class.Year, Marital.Status, Major, Next.Degree, AttendenceEvent)
* FY16Giving.Grouped vs. (FY15Giving.Grouped, ...., FY12Giving.Grouped)
* Gender vs. (Class.Year, Marital.Status, Major)
* Major vs. Next.Degree
* Class.Year vs. AttendenceEvent

###2.4.1 FY16Giving.Grouped vs. Gender

We note two interesting observations in table 1. First, there are more donations in the [$500-$200K) bracket than the [$250-$500) bracket. Furthermore, at $100 donation level or above, men consistently donate more than women do.

```{r fig.height= 3}
(t1 <- xtabs(~ Gender + FY16Giving.Grouped , data = givings))
(t1.1 <-round(t1/rowSums(t1),2))
```


###2.4.2 FY16Giving.Grouped vs. Class.Year

There are 3 key insights from table 2 below:

1. Older alumni make disproportionately larger donations (15% of the Class of 72 made $500 + donations).

2. A higher percentage of the older alumni make donations ($0 donations is only 48% for the class of 1972, versus 66% for the class of 2012). 

3. But there are more recent graduates, perhaps in part because their current addresses still valid. So even as their ratio is lower, most of the $1-$100 donations come from the class of 2012.


```{r}
(t2 <- xtabs(~ Class.Year + FY16Giving.Grouped , data = givings))
(t2.1 <- round(t2/rowSums(t2),2))
```

###2.4.3 FY16Giving.Grouped vs. Marital.Status  

Married and single people are biggest source of donations when compared to divorced or widowed alumni. We
therefore expect  Marital.Status to be a significant explanatory variable in our final model. However, there are many more married or single individuals compared to divorced or widowed alumni.

```{r}
(t3 <- xtabs(~ Marital.Status + FY16Giving.Grouped, data = givings))
(t3.1 <- round(t3/rowSums(t3),2))
```


###2.4.4 Donation level vs. Major

There are 45 majors in the dataset and as mentioned, some majors only have a single record. During model estimation, it would not be appropriate include all 45 majors individually in the model as binary variables for two reasons:

1. It will cause a curse of dimensionality that will reduce predictive power.

2. These binary variables will hold most of their records as zero and we have very little information about them.

Therefore, we've decided to group majors and after considering several ways of doing so, the method we found most appropriate was to group majors by their median donations in the last 4 years (2012-2015). Therefore, we began by calculating the average donation for each person over the last four years, and could then calculate the median donation over the last 4 years for each major. 

To optimize grouping of majors, we began with very granular cuts in the median donation (5 dollars increases). Importantly, because it is a median value, it is much less extreme than the original donation
amount (see histogram below). For half of the majors, the median donation is less than 13.75 and only 25%
of majors make median donations larger than 33.75, so the original cut-offs we assigned [0, 1), [1, 100), [100, 250), [250, 500), [500, 200000) won’t work for grouping major because most of the values will be skewed in the [1:100) category. Using the more granular cutoffs, we were able to observe the following patterns to make our final major groupings:

1. There is an extreme value at 400 in the histogram than came from an English Journalism alumnus. The English Jouralism major has a single alumnus in the dataset and this person donated $1500 in 2015 and has donated every year from 2012 to 2015. This is a caveat in our analysis because we are extrapolating from the donations made by a single person to represent the behavior of other alumni with that major. We proceeded by labeling English Jouralism as a high donation level major, but will definitely need more data to justify this point if possible.

2. [0,1) is a natural cut-off point which means no one in this major donates

3. Majors with median donations in [1,10) are showing a similar behavior in 2016 donations (70% if alumni
in these majors didn’t donate in 2016). [10-35) are showing a similar behavior in 2016 donations, with
50-60% alumni no donations, 20% donating 1-100 and about 10% donating 100-250 and 5% in following
two categories . Less than 25% of the major median are larger than 35, we can group them together.

4. We therefore decided to use the following dollar value cut-offs: [0,1),[1,10),[10,35) and 35+ to group the median donation values and we gave them the follwoing donatin category labels: “No”,“Low”,“Medium”, and “High” donation. We anticipate that that the higher the  donation level is for a major, the higher the donation these alumni will make in 2016. This is consistent with our third contingency table in this section. As the major donation_level increase from No to High, the percentage of 2016 donation in [0,1) decreases while the percentage in higher buckets like [250-500),[500-200000) increase.

Based on this analysis, we classified majors into four donation categories: “No”,“Low”,“Medium”, and “High” donation majors. We will use this variable when we estimate our model.


```{r fig.height=2.5}
givings$Last.4Year.Avg<- rowMeans(givings[c("FY12Giving","FY13Giving","FY14Giving","FY15Giving")])
Major.Index<-data.frame(aggregate(Last.4Year.Avg ~ Major, data = givings,median))
ggplot(Major.Index, aes(Last.4Year.Avg))+geom_histogram(bins=30)
givings<-merge(x = givings, y = Major.Index, by = "Major")
givings$Major.Donation.Level<-factor(cut(givings$Last.4Year.Avg.y,
                                         labels = c("NO","Low","Medium","High"),breaks=c(0,1,10,30,200000),  right = FALSE))
t18 <-xtabs(~Major.Donation.Level+FY16Giving.Grouped, data=givings)
round(t18/rowSums(t18),2)
```


###2.4.5 FY16Giving.Grouped vs. Next.Degree

A higher proportion of people with above_bachelor degree make top donations ($500 or more), compared to other groups.

```{r}
t12<-xtabs(~Adv.Deg+FY16Giving.Grouped, data= givings)
round(t12/rowSums(t12),2)
```

###2.4.6 FY16Giving.Grouped vs. AttendenceEvent

The data is inline with our expectations. Among the people who donate, there is a strong correlation between attendence and donations. In fact, most of the top donors (52 out of 59, 85%) have attended an Alumni event.

```{r}
(t4 <- xtabs(~ AttendenceEvent + FY16Giving.Grouped , data = givings))
```


###2.4.7 FY16Giving.Grouped vs. previous years' Donation levels

To analyze the relationship between FY16 donation and previous years' donations, we plotted previous years donations against the FY16 donations. The x-axis is the previous donation(FY12Giving to FY15 Giving), the y-axis is the probability of 2016 donations in each level [0, 1), [1, 100), etc.

We noticed that the probability that 2016 donations located in [0,1)("red line") decrease as previous years' donation decreases. The probability that 2016 donations fall in the in [1,100),[100,250),[250,500) range first increases with previous years' donation levels and then decreases.

In addition, the probability that the 2016 donation falls in the [500,200000) range is first unchanged with the previous years' donation but increases quickly once the previous years' level exceeds $350.
Taken together, these observations suggest that there is a clear relationship between previous years' donation levels and the target variable. However, relationship between them doesn't seem to be linear. The following contingency tables underscore this point, and clearly show how 2016 donations are highly related to previous years' donation level.

Considering the high correlation between the previous years' donation level and 2016 donation level, we wanted to carefully check whether we should include donations from all previous years in the model or whether the most recent year (2015) would suffice. For example, if an alumnus made consecutively donations in 2014 and 2015, is he more likely to donate in 2016 than an alumnus only donate in 2015?

The following plot shows the effect of 2015 donations and 2014 donations on 2016 donations. The x-axis is 2015 donations. The color shows 2014 donation while the line type shows 2016 donations (different donation levels). The contingency table shows the same information. For example, the actual probability that 2016 donation locates in [0,1) is 0.9 if the alumni neither donated in 2014 nor in 2015. This percentage decreased to 0.6 if this person donated in 2014 even if he didn't donate in 2015. The situation is similar for other year combinations. From both the plot and the contingency table, it seems we should include all of the previous years while estimating our model because every year does provide additional information to make our prediction.

```{r}
FY_analysis <- function(x,y){
groups<-aggregate(givings$FY16Giving.Grouped, by=list(givings[,x],givings[,y],givings$FY16Giving.Grouped),"length")
colnames(groups)<-c("x","y","FY16Giving.Grouped","cnt")

groups_cast<-cast(groups,x+y ~ FY16Giving.Grouped,value = 'cnt')

colnames(groups_cast)[c(1,2)]<-c("x","y")
output<- cbind(groups_cast[,c(1,2)],groups_cast[,c(3,4,5,6,7)]/rowSums(groups_cast[,c(3,4,5,6,7)],na.rm = TRUE))
output[is.na(output)] <- 0
return(output)}
output<-FY_analysis("FY15Giving.Grouped","FY14Giving.Grouped")
melt_output<-melt(output,id=c("x","y"))
levels(melt_output[,1])<- c(0,1,100,250,500)
melt_output[,1]<-as.numeric(levels(melt_output[,1]))[melt_output[,1]]

ggplot(melt_output, aes(x,value,colour=y,linetype=variable))+geom_line(lwd=1)+theme_bw()+ggtitle(paste("FY15Giving.Grouped"," vs ","FY14Giving.Grouped"))+ylab("Actual probability")+scale_linetype_manual(values=c("solid","twodash","longdash","12345678","dotted"))+scale_color_manual(values=c("red","orange","blue","green","purple"))+guides(color=guide_legend(title="FY16Giving.Grouped"),linetype=guide_legend(title="FY16Giving.Grouped"))


colnames(output)[1:2]<-c("FY15Giving.Grouped","FY14Giving.Grouped")
head(cbind(output[,1:2],round(output[,3:7],2)),4)
```



```{r}
#xtabs(~ FY16Giving.Grouped + FY12Giving.Grouped, data = givings)
#xtabs(~ FY16Giving.Grouped + FY13Giving.Grouped, data = givings)
#xtabs(~ FY16Giving.Grouped + FY14Giving.Grouped, data = givings)
#xtabs(~ FY16Giving.Grouped + FY15Giving.Grouped, data = givings)
```


###2.4.8 Gender vs. Class.Year

As expected, over the years, the gender ratio converges towards a gender neutral 50%, but in the earlier years males were a higher percentage of the sample. It is also worth noting that there is an unexpected change in the ratio for the class of 2002. We will explore the Gender:Class.Year interaction in the next section of the EDA.

```{r}
t6 <- xtabs(~ Class.Year + Gender, data = givings)
t(t6)
t(round(t6/rowSums(t6),2))
```


###2.4.9 Gender vs. Marital.Status

We previously observed that that Married and Single alumni were more likely to donate than divorced or widowed alumni, however, the vast majority of sample consists of Married and Single alumni. We anticipate that this will weaken the predictive power of the Marital.Status variable. We will create a simpler variable (Married? yes or no) to improve the parsimony of the final model. We also note here that strong skew in widow ratio can be explained by the life expectancy differences between men and women.

```{r}
t7 <- xtabs(~ Marital.Status + Gender, data = givings)
t(t7)
t(round(t7/rowSums(t7),2))
givings$Married <- factor(ifelse(givings$Marital.Status == "Married", TRUE,FALSE))
```


###2.4.10 Gender vs. Major

Here, we explored the relationship between high/medium/low donation major groups and gender

We had already established that among the high level donors, men had a higher ratio than women. We now
conclude that this is also reflected for Majors. Majors that on average had lower previous year donations had a higher percentage of females than males and majors that have the highest donation levels have more males than females.


```{r}
t8 <- xtabs(~  Major.Donation.Level + Gender, data = givings)
t(t8)
t(round(t8/rowSums(t8),2))
```


###2.4.11 Major vs. Next.Degree (Adv.Deg)

We examined the distribution of alumni with or without next degrees based on the donation level of their major. We found that alumni who did majors characterized by high donation levels most frequently either did not have a second degree, or had a next degree that was above the bachelors level. Surprisingly, alumni who did a second bachelor's degree were the least likely to give at any donation level.

```{r}
t99 <- xtabs(~ Major.Donation.Level + Adv.Deg, data = givings)
t(t99)
t(round(t99/rowSums(t99),2))
```


###2.4.12 Major vs. AttendenceEvent 

Based on our segmentation of each Major's donation level, we find that there is a linear trend between Attendance and Major Donation Level, namely that as the donation level increased, so did the percentage of event attendees. This is not surprising considering we saw earlier that Attendance is correlated with donation level.

```{r}
t9 <- xtabs(~ Major.Donation.Level + AttendenceEvent, data = givings)
t(t9)
t(round(t9/rowSums(t9),2))
```


###2.4.13 Class.Year vs. AttendenceEvent

It is remarkable that graduates from 1972 have the same (~60%) attendance rate as the class of 2012. Of note, the Class of 2002 has an usual spike in attendance, but we cannot explain it with the information available to us.

```{r}
t10 <- xtabs(~ Class.Year + AttendenceEvent, data = givings)
t(t10)
t(round(t10/rowSums(t10),2))
```


##2.5 Interactions

### 2.5.1 Gender, Class.Year, 2016 Donations

```{r}
xtabs(~ Class.Year + FY16Giving.Grouped + Gender, data = givings)

```

The difference in top donations (above $500) can be explained by male / female ratio. For example, 6 women made $500+ donations from the class of 72, vs. 10 man. However, their ratio (6/10) is not too far from the female/male ratio (0.56) for the class of FY72. We are not anticipating a strong interaction between Gender, Class.Year, and 2016 Donation levels.  

### 2.5.2 Gender, Major, 2016 donations
```{r}
xtabs(~ Major.Donation.Level + FY16Giving.Grouped + Gender, data = givings)
```

Looking at the top donors from Majors that have usually high donation levels, we see that they are more likely to be male. We know that top donor Majors (Major.Donation.Level == High) have a 3-to-1 Male/Female ratio. But we observe a 7-to-0 ration for Male/Female distribution for donors who have donated above $500 and are from top donor majors. **So we we believe there may be an interaction between Gender, Major, 2016 donations**



##3 Statistical Modeling 

#Summary


The final model we estimated is:

$$ logit(\hat{P}(Y\leq j)) = \hat{\beta}_{j0} + 0.07Married + 0.27Attended + 1.21FY15Giving \ ... + 2.10FY12Giving.Grouped $$
where $\hat{beta}_{10} = 2.73, \hat{beta}_{20} = 4.38, \hat{beta}_{30} = 6.82, \hat{beta}_{40} = 8.21$

We found that past years giving is the key predictor for 2016 contributions. There are increasing Odds of donation for alumni who previously donated in high donation brackets. This trend applies to all previous Fiscal Years, but more so to the more recent donation years. However, to as lesser extent Gender, being in high donation Major category, and attendance of events increased the odds of 2016 contribution.

#Modeling Process

Based on our exploratory data analysis we decided to include most of the variables provided in the dataset as explanatory variables while estimating our model. These included all years of FYGiving.Grouped (2012-2015), our categorical variable of Married, AttendenceEvent, Major.Donation.Level, Adv.Deg,Gender, and as described below we originally included Class.Year, but removed it from the model due to poor model fit.


#Ordinal versus Nominal Multinomial model?

We noted that there was a natural order to ....

When reviewing each Independent Variable, we found that their odds ratios were relatively similar across the different levels of the response variable FY16Giving.Grouped. This provided additional justification for estimating an Ordinal model.

```{r}
c1 <- xtabs(~Gender+FY16Giving.Grouped, data=givings)
c2 <- xtabs(~Marital.Status+FY16Giving.Grouped, data=givings)
c3 <- xtabs(~Class.Year+FY16Giving.Grouped, data=givings)
c4 <- xtabs(~Major.Donation.Level+FY16Giving.Grouped, data=givings)
c5 <- xtabs(~AttendenceEvent+FY16Giving.Grouped, data=givings)

odds_ratio <- function(r1){
  df1 <- as.data.frame.matrix(r1)
  n <- dim(df1)[1]
  len <- dim(df1)[2]
  odds = data.frame(matrix(0, n, len-1))
  colnames(odds) <- colnames(df1)[1:len-1]

  for( i in seq(1,len-1)){
    if(i==1) {
      lowerp <- df1[,1]
    }else {
      lowerp <- rowSums(df1[,1:i])
    }
    if(i==len-1) {
      upperp <- df1[,len]
    } else {
      upperp<-rowSums(df1[,(i+1):len])
    }
    odds[,i] <- lowerp/upperp
  }
  round(odds,2)

  oratio <- data.frame(matrix(0, n-1, len-1),row.names = rownames(df1)[2:n])
  colnames(oratio)<- colnames(odds)
  for(j in seq(1,n-1)) oratio[j,] <- odds[j+1,]/odds[j,]
  return(round(oratio,2))
}
#Gender
odds_ratio(c1)
#Marital Status
odds_ratio(c2)
#Graduating class
odds_ratio(c3)
#Major
odds_ratio(c4)
#Addent Event
odds_ratio(c5)
```

In analysis we do not show for brevity, we identified that FY15Giving.Grouped had the highest in-model explanatory power of all explanatory variables by themselves (the lowest AIC and it was statistically significant). Thus, we chose model_B1a as our "base model" to use during hypothesis testing.


```{r}
model_B1a <- clm(formula= FY16Giving.Grouped ~ FY15Giving.Grouped, data = givings, link="logit")
Anova(model_B1a)
AIC(model_B1a)
```

All the models we estimated are summarized in table below (*NB that the table displays Wald SE
and p values based on the Wald Test, however our Likelihood Ratio tests come to the same conclusion as the
Wald Tests).


```{r}
#our original model based on including all variables that seemed related to FY16Giving.Grouped in EDA
model_B1g <- clm(formula = FY16Giving.Grouped ~ Class.Year + Married + AttendenceEvent + FY15Giving.Grouped + Major.Donation.Level + Adv.Deg+Gender+FY14Giving.Grouped + FY13Giving.Grouped + FY12Giving.Grouped,data =givings, link ="logit")
#Anova(model_B1g)
anova(model_B1a, model_B1g)
```

Our null hypothesis, Ho, is that:

$\beta_{Class.Year}=\beta_{Married}=\beta_{AttendenceEvent}=\beta_{Major.Donation.Level}=\beta_{Adv.Deg}=\beta_{Gender}=\beta_{FY14Giving.Grouped}=\beta_{FY13Giving.Grouped}=\beta_{FY12Giving.Grouped} = 0$

Our alternative hypothesis is that:

$\beta_{Class.Year}\neq\beta_{Married}\neq\beta_{AttendenceEvent}\neq\beta_{Major.Donation.Level}\neq\beta_{Adv.Deg}\neq\beta_{Gender}\neq\beta_{FY14Giving.Grouped}\neq\beta_{FY13Giving.Grouped}\neq\beta_{FY12Giving.Grouped} \neq0$


By comparing the FY15Giving.Grouped-only model (model1.Ho) to the model with all years of FYGiving.Grouped (2012-2015), our categorical variable of Married, AttendenceEvent, Major.Donation.Level, Adv.Deg,Gender, and Class.Year (model1.Ha) using the Likelihood Ratio test, we found that adding the additional variables besides FY15Giving.Grouped contributed to the model - ie. there is strong support to reject the null hypothesis.

Because there is good evidence that our addition of 9 explanatory variables (model.Ha) to our base model (model.Ho) made a singificant contribution to the model, and because the AIC decreased when we added the explanatory variables as suggested by our EDA (model.Ho = 1705.7, model.Ha = 1560.5), we proceeded with evaluating model fit.

We are aware of the strong correlation between donation levels from 2012 to 2016. However, we've decided to keep them as well for two reasons: 1. They are contributing to the model prediction significnatly(shown by the ANOVA test). 2 The predicted coefficient looks consistent with our observation in EDA. 3 Although a lot of categorie variables are not signficant in Wald test because of inflated standard error, it doesn't necessary negatively the final prediction.

We found that the Class.Year is neither significant in the model nor reports a correct coefficient. On the top of that, in our EDA it is clear that earlier (specially for 1972) graduates is more likely to donate high while the coefficient is showing the opposite, which is very misleading for our model interpretation. The reason for the problem is that Class.Year highly correlated with all the previous donation variable, which increases the standard error (leading to unsignficant test) and incorrect coefficient. Finally we find it may necessarily to keep Class.Year in this model since it is strongly correlated with all previous donation levels and doesn't provide additional information(the model with or without this variable is not significantly different from each other).

Considering the poor fit of model.Ha, we removed Class.Year and re-evaluated the model we estimated with the remaining explanatory variables.

Our null hypothesis now was that:

$\beta_{Married}=\beta_{AttendenceEvent}=\beta_{Major.Donation.Level}=\beta_{Adv.Deg}=\beta_{Gender}=\beta_{FY14Giving.Grouped}=\beta_{FY13Giving.Grouped}=\beta_{FY12Giving.Grouped} = 0$

```{r}
#remove Class.Year as including it causes poor predictions due to high collinearity with FYGiving.Grouped variables
model_B1j <- clm(formula = FY16Giving.Grouped ~ Married + AttendenceEvent + FY15Giving.Grouped + Major.Donation.Level+Adv.Deg+Gender+FY14Giving.Grouped+FY13Giving.Grouped+FY12Giving.Grouped,data =givings, link ="logit")
summary(model_B1j)
Anova(model_B1j)
anova(model_B1j,model_B1a)
```



```{r}
conf.beta <- confint(object = model_B1j , level = 0.95)
conf.beta <- as.data.frame.matrix(conf.beta)
conf.beta[,3]<-model_B1j$coefficients[5:28];colnames(conf.beta)[3] = "estimated"
round(conf.beta,2)
round(exp(conf.beta),2)
```






# ORDINAL REGRESSION

The estimated model is:

$$ logit(\hat{P}(Y\leq j)) = \hat{\beta}_{j0} + 0.07Married + 0.27Attended + 1.21FY15Giving \ ... + 2.10FY12Giving.Grouped $$

where $\hat{beta}_{10} = 2.73, \hat{beta}_{20} = 4.38, \hat{beta}_{30} = 6.82, \hat{beta}_{40} = 8.21$

To fit the model, we use the model using the $clm$ function below:

```{r}
#model_B1j <- clm(formula = FY16Giving.Grouped ~ Married + Adv.Deg + Gender + AttendenceEvent  + FY15Giving.Grouped + FY14Giving.Grouped +FY13Giving.Grouped + FY12Giving.Grouped, data =givings, link ="logit")
```

# INDEPENDENCE

We test for independence among the variables using the Anova function, with:

$$H_0 : \beta_2 = \beta_3 = ... \beta_22 = 0$$
$$H_a : any \ \beta_i \neq 0$$

From the test, we find that there is strong evidence that association exists for all variables except for AttendenceEvent and Major.Donor.Level.

```{r}
Anova(model_B1j)
```

# CONFIDENCE INTERVAL

We derived the confidence intervals for the Odds Ratios to determine the significance and impact of the variables within our model:

```{r}
conf.beta <- confint(object = model_B1j , level = 0.95)
conf.beta <- as.data.frame.matrix(conf.beta)
conf.beta[,3]<-model_B1j$coefficients[5:28];colnames(conf.beta)[3] = "estimated"
round(conf.beta,2)
round(exp(conf.beta),2)
```

The table above details the range of Odds Ratios for any given variable compared with its base, holding all other variables constant. For example, with 95% confidence* the odds of an alumni donating above a given donation bracket is .72 to 1.36 times when they are married than when they are not. The most notable trend is among the previous Fiscal Year donations, as there are increasing Odds for a donation for alumni who donated in higher donation brackets in previous years. Specifically, we see that if an alumni donated in the highest bracket (500-200000) in FY15, his/her odds of donating above a given bracket are 39 to 417 times compared to if he/she did not donate in FY15. This trend holds true for all previous Fiscal Years, although to a lesser extent the less recent the donation year.

A few addition variable interpretations:

- The odds of donating above a certain bracket are between .51 to 1.26 times as large when the alumni had a bachelor or equivalent degree versus no advanced degree. This range is reduced to .34 to .89 times as large when the alumni had above a bachelor degree compared to no advanced degree.
- The odds of donating above a certain bracket are between .85 to 1.54 times as large when the alumni was a male compared to female.
- The odds of donating above a certain bracket are between .94 and 1.80 times as large when the alumni attended at least one event compared to not attending any events.
- The odds of donating above a certain bracket are between .74 to 1.40 times as large when the alumni majored in what we categorized as a 'High Donor' major compared to other majors.

We notice that many of the variables include 1 within the 95% confidence interval. This suggests that there is not sufficient evidence to indicate that these variables increase the odds of donating. However, through our EDA and logical assessment, many of these variables can be key contributers and also add information to other variables in the model, thus we chose to keep them.

*95% confidence is defined in the frequentist perspective, meaning given 100 samples drawn from the same population, we can expect 95 of the samples to have a 95% confidence interval that contains the true population parameter.

#MODEL FIT

```{r}
#explanatory variables
var.list <- c("Married","AttendenceEvent","FY15Giving.Grouped","Major.Donation.Level","Adv.Deg","Gender","FY14Giving.Grouped","FY13Giving.Grouped","FY12Giving.Grouped")
pred.list<-c("[0-1)","[1-100)","[100-250)","[250-500)","[500-200000)" )
newData <- unique(givings[,var.list])
predicts<-cbind(newData, predict(model_B1j, newdata=newData)$fit)

predicts<- merge(predicts,givings,by=var.list)
FY15.pred.mean <- ddply(predicts, "FY15Giving.Grouped", function(x) colMeans(x[pred.list]))
FY15.pred.mean <- cbind(FY15.pred.mean[,1],FY15.pred.mean[,2:6]/rowSums(FY15.pred.mean[,2:6]))
colnames(FY15.pred.mean)[1]<-"FY15Giving.Grouped"
FY15.actual<-xtabs(~FY15Giving.Grouped+FY16Giving.Grouped,givings)
FY15.actual<-as.data.frame.matrix(FY15.actual/rowSums(FY15.actual))
FY15.actual<-FY_actual("FY15Giving.Grouped")

melt.method<- function(df,x,type){
melt_predicts<-melt(df,id=x)
melt_predicts$type<-type
levels(melt_predicts[,1])<- c(0,1,100,250,500)
melt_predicts[,1]<-as.numeric(levels(melt_predicts[,1]))[melt_predicts[,1]]
return(melt_predicts)
}
FY15.actual<- melt.method(FY15.actual,"FY15Giving.Grouped","actual")
FY15.pred.mean<- melt.method(FY15.pred.mean,"FY15Giving.Grouped","predicted")
FY15.all<-rbind(FY15.actual,FY15.pred.mean)
plots3<-list()
plots3[[1]]<-ggplot(FY15.all[FY15.all$variable %in% c("[0-1)","[500-200000)"),], aes(FY15Giving.Grouped,value,colour=type,linetype=variable))+geom_line(lwd=1.5)+theme_bw()+ggtitle(paste("Predicted"," vs ","actual"))+ylab("probability")

plots3[[2]]<-ggplot(FY15.all[FY15.all$variable %in% c("[1-100)","[100-250)","[250-500)"),], aes(FY15Giving.Grouped,value,colour=type,linetype=variable))+geom_line(lwd=1.5)+theme_bw()+ggtitle(paste("Predicted"," vs ","actual"))+ylab("probability")
```


```{r fig.width=15, fig.height=6, echo=FALSE}
plot_grid(plots3[[1]],plots3[[2]],ncol = 2)
```



##4 Final Remarks


Through our analysis we've found multiple variables in the data that correlate with, and help predict, donation levels from alumni. Although our model fits the data well, we've identified a few areas that should be noted:

1. Our model leverages previous years' donations as a primary predictor of the level of donation for the current year. For the majority of alumni, this works well as we have data from previous years. For new alumni, however, this variable will be missing for a certain number of years. We have found, however, that the further back the donation data goes, the less predictive power it provides, thus this is well mitigated. It would still be beneficial to ensure we have as much historical data on alumni donation as possible, so as to make full use of our model.

2. There are certain subcategories of variables (e.g. Widowed in the Marital Status variable, certain majors in the Major variable, etc.) that have very little data. For our analysis, we've elected to group these variables (as in the Major.Grouped specification), which works well, but it is likely that having additional observations in these subcategories would help fit a better model. We understand this is not a controllable input, but we are highlighting it here for awareness.

3. Comparing the predictions from our model with the actual donation levels in FY16, we find that our predictions are far more accurate for the lowest and highest tiers of donations. This is likely because there is greater variability in the intermediate level, whereas the relationships are much stronger as you look at the extremes. In our scenario, this actually works well because a large percentage of the total donations are weighted toward the higher donation tiers, which mean that accurately predicting alumni in these tiers are more important than other tiers. Additionally, accurately predicting those who will not donate is also helpful so the administration can allocate their outreach accordingly.

4. We approached this analysis with an in-sample EDA and model assessment as we wanted to leverage as much of the data in our analysis as possible, knowing that additional data from next year would be relatively easy to attain and test on. Given a larger sample, we would be more comfortable splitting the data in to a training and testing sample, especially given the small subcategories from the variables mentioned in the previous point.

As with most analyses, more data can lead to better models, but we've analyzed the existing data set and developed a model which we think fits the data and will predict the level of donations well. We recommend the administration leverage this model to predict FY17 Donation levels and check it against the actuals to confirm the model's accuracy, particularly on the highest level donations.

We do, however, note that fundraising professionals typically make use of datasets that include some measure of wealth. Rather than use proxies for wealth (like Next Degree and high donor Major category for example), we recommend that the Univeristy obtain actual income level information for alumni. If this is not possible, there are other ways fundraisers commonly use to estimate wealth that include estimating potential donors' home values using donor's home addresses - people who own $2 million dollars or more in real estate are 17 times more likely donate to a nonprofit than the average person (www.bidpal.com/identifying-major-donors-top-strategies-tools). By identifying wealthy alumni, the University could invite them to "VIP University Events" and cultivate relationships with these donors that tend to result in larger donations.










