---
title: "W271 Section 3 Lab 2"
author: "Kiersten Henderson, Jill Zhang, Hoang Phan, Daghan Altas"
date: "10/11/2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, warning=FALSE}
knitr::opts_chunk$set(cache=TRUE)
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

#1 Introduction  
An introduction to the project, which should include a concise summary of the key results as well as techniques you used in your final model.

#2 Exploratory Data Analysis
```{r}
library(dplyr)
library(ggplot2)
library(Hmisc)
library(GGally)
library(data.table)
library(stargazer)

if (dir.exists("XXXX")) {
  ## For Kiersten
  setwd("XXXX")
} else if (dir.exists("YYY")) {
  ## For Jill
  setwd("YYY")
} else if (dir.exists("ZZZ")) {
  ## For Hoang
  setwd("ZZZ")
} else {
  ## For Daghan
  setwd("/Users/daghanaltas/Hacking/Berkeley/W271/Labs/w271_lab2/")
}
 
df <- read.csv("./lab2data.csv")
dt <- data.table(df)
head(dt)
describe(dt)
```
## Todo

* Missing values (NAs)?
* Create factor variables for Gender, Marital Status, Class year, Major, Next.Degree, 


#3 Statistical Modeling  
Questions to answer: 

* Start the section summarizing the key results - what variables, if any, are the key predictors of the year 2016 contribution? 
* What are the key techniques you have experimented? 
* What method did you use in your final model? 
* How did you choose the final model? 
* What model performance criteria did you use to choose the final model? 
* What statistical inference did you perform? Explain them. 
* Comment on statistical significance vs. economic significance.

#4 Final Remarks  
Questions to answer:  

* What are your departing thoughts? 
* What are the strengths and weaknesses in your analysis? 
* Should the administration trust your result? 
* Are there subsample in your sample that your model did a bad job in predicting their contribution behavior?
    + If so, why? 
* Are there other “things”, a wish list, that you think can be used to improve your model?
    + If so, what are they?
* Perhaps you can make a suggestion to the administration to collect those information in the future.






```{r}
```